A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-09 paper.tex
More meat in the Genome ARGs section

\section*{Genome ARGs}
% There is no fundamental reason, however, that the data structure we use
% to describe patterns of genetic inheritance must be in terms
% of specific causal events. As demonstrated in subsequent sections,
% there are significant advantages to using
% a slightly different formulation, which
% we call a ``genome ARG'' (gARG).
In this section we define a concrete definition of an ARG in terms
of particular genomes and their patterns of inheritance, which
to avoid confusion, we refer to as a ``genome ARG``, or gARG.
(We contrast this with the classical definition derived from the
coalescent process in later sections.)
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
Then, a gARG is a graph in which nodes represent
monoploid genomes  and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance).
% To recover the local tree for a genomic
% position $x$, we traverse the graph rootwards from the leaves
% (as in an eARG) and at a particular node, we follow the outbound
% edge associated with the genomic interval containing $x$.
% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
Fig.~\ref{fig-arg-in-pedigree}
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
In this diploid setting
(there is no limitation on mating system, ploidy, or age structure)
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
% NOTE: does this example really help? Rather than spelling out what
% we can already see, perhaps we should point out the non-obvious
% thing that we have decoupled from "events"
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.
Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
etc) can be expressed without difficulty using this encoding.

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An Ancestral Recombination Graph defined in terms of genomes embedded
in a pedigree. (A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal (left) and maternal (right) genomes
labelled \textsf{a} to \textsf{p}. Black lines show transmission paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Two recombination breakpoints, at positions 2 and 7 mean that genomes \textsf{a}
and \textsf{c} are independent recombinants of the paternal genomes \textsf{e}
% Gregor: there is some overlap between a and c still (the middle region inherited
% from f) so, a and c are not fully independent!
and \textsf{f}: in these cases the edge annotations have been shown, indicating the
region of genome transmitted. Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker; this shows that all
genomic regions find a common ancestor in genome \textsf{n} in the oldest generation.
(B) The corresponding genome ARG, where there is a single type of node (representing
a genome) along with annotations on all edges; annotations showing partial
transmission are highlighted in bold. Recombination (e.g.,
the cause of the breakpoint at position 7)
is associated with two parent genomes (\textsf{e} and
\textsf{f}) and a child genome (the recombinant, \textsf{c}).
(C) The edge-annotated genome ARG in (B) fully defines the
local relationships along the genome, as shown in the local
(``marginal'') trees.
}
\end{figure}

Formally, we can define a gARG as a set of edges $E$, where each
element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
also phrase this in terms of the ancestral material of a sample---see
the next section).
% To recover the local tree for a genomic
% position $x$, we traverse the graph rootwards from the leaves
% (as in an eARG) and at a particular node, we follow the outbound
% edge associated with the genomic interval containing $x$.
To recover the local tree at genome coordinate $x$ (the most
fundamental operation required for an ARG data structure; see
Section XXX),
we traverse the graph rootwards from the leaves.
Starting with the sample nodes $S$ we traverse
rootwards through the graph from child to parent. At each node $u$, we find an
edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$
and move to the next node $p$.

[COMMENT ON FIG 1C, and unary nodes in the trees]

>>>>

\section*{Genome ARGs}
% There is no fundamental reason, however, that the data structure we use
% to describe patterns of genetic inheritance must be in terms
% of specific causal events. As demonstrated in subsequent sections,
% there are significant advantages to using
% a slightly different formulation, which
% we call a ``genome ARG'' (gARG).
In this section we define a concrete definition of an ARG in terms
of particular genomes and their patterns of inheritance, which
to avoid confusion, we refer to as a ``genome ARG``, or gARG.
(We contrast this with the classical definition derived from the
coalescent process in later sections.)
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
Then, a gARG is a graph in which nodes represent
monoploid genomes  and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance).
% To recover the local tree for a genomic
% position $x$, we traverse the graph rootwards from the leaves
% (as in an eARG) and at a particular node, we follow the outbound
% edge associated with the genomic interval containing $x$.
% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
Fig.~\ref{fig-arg-in-pedigree}
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
In this diploid example
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
% NOTE: does this example really help? Rather than spelling out what
% we can already see, perhaps we should point out the non-obvious
% thing that we have decoupled from "events"
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An example gARG embedded in a pedigree.
(A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal  and maternal  genomes
labelled \textsf{a} to \textsf{p}. Black lines show transmission paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Genomes \textsf{a} and \textsf{c} are independent recombinants of
the paternal genomes \textsf{e}
% Gregor: there is some overlap between a and c still (the middle region inherited
% from f) so, a and c are not fully independent!
and \textsf{f}, and regions of genome transmitted are shown.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding genome ARG along with annotations on all edges;
annotations showing partial
genome transmission are highlighted in bold. Recombination (e.g.,
the cause of the breakpoint at position 7)
is associated with two parent genomes (\textsf{e} and
\textsf{f}) and a child genome (the recombinant, \textsf{c}).
(C) The edge-annotated genome ARG in (B) fully defines the
local relationships along the genome, as shown in the local
(``marginal'') trees.
}
\end{figure}

Formally, we can define a gARG as a set of nodes $N$ and edges $E$, where each
element of $E$ is a tuple $(c, p, I)$ such that
$c\in N$ is the child node, $p\in N$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
also phrase this in terms of the ancestral material of a sample---see
the next section).
% To recover the local tree for a genomic
% position $x$, we traverse the graph rootwards from the leaves
% (as in an eARG) and at a particular node, we follow the outbound
% edge associated with the genomic interval containing $x$.
To recover the local tree at genome coordinate $x$ (the most
fundamental operation required for an ARG data structure; see
Section XXX),
we traverse the graph rootwards from the leaves.
Starting with the sample nodes $S$ we traverse
rootwards through the graph from child to parent. At each node $u$, we find an
edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$
and move to the next node $p$. The efficiency of this process is discussed
in Section XXX).

There are some interesting consequences to these definitions, exemplified
in Figure~\ref{fig-arg-in-pedigree}. Firstly,
there is
is no limitation on mating system, ploidy, or age structure. By making the
basic unit of analysis an individual's \emph{genomes} (i.e., two per individuals
for diploids), any pattern of inheritance among mixed ploidy can be
accommodated.
Secondly, because we only record the \emph{intervals} of inheritance
between genomes (and not the causal \emph{processes}),
arbitrary patterns of genetic inheritance such as gene conversions,
multiple recombinations and so on can be directly expressed.
Thirdly, ``local`` trees that we generate for a particular position on the
genome are not explicitly concerned with ``coalescence'' (nodes at which
one or more samples find a common ancestor); for example, node $e$
is unary in the left-most tree of Figure~\ref{fig-arg-in-pedigree},
denoting the fact that $e$ the ancestor of $a$ over the interval $(0, 2]$,
but no coalescence occurs in this interval (see section XXX for more
discussion). Fourth, recombination is modelled in terms of
its outcomes in terms of inheritance intervals among parental genomes;
thus, we will have a child node (the recombinant; e.g. node $a$
Figure~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
paternal genomes which this child has inherited segments from.
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the evolutionary processes of genetic inheritance and
recombination, and is agnostic to the details of those processes.
